Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-05-30 analysis/paper/paper2.Rmd
Final writing and first edits

---
title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
subtitle: "Early Iron Age Cemeteries at Knossos"
author:
  - Dominic Pollard, Institute of Archaeology, University College London
  - dominic.pollard.16@ucl.ac.uk
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
bibliography: Lib.bib
csl: "../templates/world-archaeology.csl" # Insert path for the bib-style

abstract: |
  The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site between the end of the Bronze Age and the emergence of the Archaic *polis*. However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken. This paper does just that, and in so doing identifies, and attempts to explain, a pattern of increased burial rates among a few select tombs. This paper demonstrates the potential for quantitative analyses to contribute towards nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age. It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use.
keywords: |
  Early Iron Age; Knossos; Burial; Quantitative; Open-Access; Reproducibility

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.retina = 2,
  comment = "#>",
  fig.path = "../figures/",
  dev="jpg"
)

@@

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971] and culminating in the rejection by many scholars of the image of a society bereft of its former palatial glory. Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.

But more importantly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period. This is reflected too in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra, where such sites are numerous. This has, unfortunately, engendered a certain disregard for settlement characterised by quite different topographies, especially in the central region of the island. There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses for elucidating social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research [@Bevan2013; @Knappett2011a]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b]. But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, but a general impression of un-patterned multiplicity.

I suggest that the application of a range of quantitative and statistical methods may provide new insight into the social behaviours underlying the mortuary assemblages. With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries. In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287]. Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace. Indeed, Alexiou [@1950] thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.

@@

## The Early Iron Age Cemeteries
```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
knitr::include_graphics("../figures/Knossos Tombs Map.jpg")
```

Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997 @Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed and often isolated that they provide little scope for systematic quantitative analysis.

@@

## The Fortetsa and the Knossos North Cemeteries

The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957, xi). These tombs cluster in three main groups. The largest of these, comprising nine tombs, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west. To the north lies a cluster of five tombs and, to the south, another of three. Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2], are now considered to belong the Knossos North Cemetery (KNC).

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the KNC. Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb. They are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978. Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration in the whole Knossos area [@Coldstream1996a, 53-55].

The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012, 47-8]. The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts. 200, 201 and 202 are three chambers of a single pit tomb) in the KNC [@Coldstream1994, 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts. L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb [@Antoniadis2012, 69].

## Chronological developments in the cemeteries

The initial foundation of the cemeteries has received much attention. It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), though there is a possibility that this is a classificatory artefact [@Hallager2010]. Instances of BA tombs reused the EIA are known (e.g. the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers. comm.).

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment. The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, and evinces a melding of Attic geometric motifs, experimental patterns possibly deriving from Near Eastern metalwork, and iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4). Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T. 107; see @Coldstream1984b; -@Coldstream1994, 112–3). Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-5], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.

@@

Catling’s discussion of the Tomb 200+ complex is a notable example. Tomb 201 contained the remains of at least two adults, and possibly a child, a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi. All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials. Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing precious artefacts and stories of distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].

Complementing the theme of heroes abroad has been a similar interest in foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the burial place of an immigrant Phoenician craftsman. Kotsonas has revisited this attribution, suggesting that the tholos is more likely that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9]. While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material. Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies. Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, the implication being that a hierarchy existed, with unequal access to such prestigious goods. Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore. The present paper seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel, by far the largest portion of all known finds. Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016). In total, this database comprises just over 4540 individual artefacts.

@@

## Reproducibility and data-sharing

An ever-growing number of archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty and further investigation. It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles - and excuses - preventing the full and free sharing of both data and workflows. In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

This is not the place for a detailed investigation into the lack of reproducible research in archaeology. Bu this article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete. Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present data such that they could be manually digitsed, is less pardonable.

There are, of course, complications to the sharing of data and methods. There are real and valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant investment of time demanded by the preparation of open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis. Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R. This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (NB a URL will be included pending acceptane of the article). The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data. It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

## Estimating interments

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)). In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).

```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters. The straight line represents the mean estimate based on these three values."}

@@

##Aoristic approaches to dating

A central concern of the present study was how to deal with the temporal uncertainty implicit in the assigning of individual vessels to stylistically-defined periods. A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data through means which normally rely on value- and not range-based data inputs. A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper when the division of vessels into centuries renders the inaccuracies of such a method less significant.

```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}

@@

A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis" [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades. These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures. The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time."}

@@

Table: Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries, through each of the centuries they were in use, as well as the number of tombs from which these burials came, allowing calculation of the changing rate of burial through time.

>>>>

---
title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
subtitle: "Early Iron Age Cemeteries at Knossos"
author:
  - Dominic Pollard, Institute of Archaeology, University College London
  - dominic.pollard.16@ucl.ac.uk
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/world-archaeology.csl" # Insert path for the bib-style

abstract: 
   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*. However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken. This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries. Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and goes on to consider the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups. This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age. It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use"

---
# Keywords: 
Early Iron Age; Knossos; Burial; Quantitative; Open-Access; Reproducibility

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.retina = 2,
  comment = "#>",
  fig.path = "../figures/",
  dev="png"
)

@@

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.

Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period. This is reflected in survey work, which has focussed extensively around the Bay of Mirabello, where such sites are numerous [@Haggis2005; @Hayden2005; @Watrous2012]. This has, unfortunately, engendered a certain disregard for settlement characterised by different topographies, especially in the centre of the island. There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been studied with regard to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b]. But a lack of quantitative analyses, or systematic comparisons between the cemeteries, has left us with many intuited trends, some exceptional tombs, but a general impression of un-patterned multiplicity.

With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis seeks to address this present lack. In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287]. Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].

@@

## The Early Iron Age Cemeteries

```{r map, fig.cap="goes around here"}
knitr::include_graphics("../figures/Knossos Tombs Map.jpg")
```

Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997 @Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed and often isolated that they provide little scope for systematic quantitative analysis.

@@

## The Fortetsa and Knossos North Cemeteries

The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi). These tombs cluster in three main groups. The largest of these, comprising nine tombs, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west. To the north lies a cluster of five tombs and, to the south, another of three. Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the KNC. Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb. They are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978 [@Coldstream1996a: 53-55].

The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-8]. The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials and [@Antoniadis2012: 69].

## Chronological Developments in the Cemeteries

The initial foundation of the cemeteries has received much attention. It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582], though there is a possibility that this is a classificatory artefact [@Hallager2010]. Instances of BA tombs reused in the EIA are known (e.g. the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers. comm.).

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment. The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957, 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b, 93-94]. Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994, 112–13]. Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-15], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various EIA burying groups.

@@

Catling’s discussion of the Tomb 200+ complex is a notable example. Tomb 201 contained the remains of at least two adults, and possibly a child, a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi. All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials. Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].

Complementing the theme of heroes abroad has been a similar interest in foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman. Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9]. While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material. Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore. The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe (2016). In total, this database comprises just over 4540 individual artefacts.

@@

## Reproducibility and Data-sharing

Evermore archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty. It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article]. In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete. Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present data such that they could be manually digitsed, is less pardonable.

There are, of course, complications to the sharing of data and methods. There are valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis. Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R. This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**). The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data. It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

## Estimating Interments

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)). In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

##Aoristic Approaches to Dating

A central concern of the present study was how to deal with the temporal uncertainty implicit in the assigning of individual vessels to stylistically-defined periods. A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data through means which normally require values and not ranges. A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper when the division of vessels into centuries renders the inaccuracies of such a method less significant.

```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis" [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades. These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures. The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

*Table 1 goes around here*
